# Handling data

Base R refers to functions that come with an installation of `R`, not those that are in packages that are must be installed. We will learn how to handle data both using the base R functions, and using the package `dplyr`.

## Base R

### Useful functions

-   `max()` outputs the maximum value in a vector (recall, a vector is a list of numbers, e.g., `c(0,2,5)`), matrix, or data frame;

-   `min()` outputs the minimum value in a vector, matrix, or data frame;

-   `length()` outputs the total number of elements, in a vector, matrix, or data frame;

-   `sum()` adds or takes the sum of all values in a vector, matrix, or data frame;

-   `mean()` calculates the mean of all values in a vector or matrix;

-   `colMeans()` calculates the mean of a column in a matrix or data frame.

- `head()` prints the first 6 rows of the argument to your console. This can be helpful if you want to check your data without having to sort through a long output.

- `tail()` similar to `head()` but prints the last 6 rows of the argument.

**TRY IT!** In the `Console` do some tests on these functions by creating a test matrix and a test data frame:

```{r}
test = data.frame(col1 = c(0,2), col2 = c(5,6))
test2 = matrix(c(2,3),c(5,1),nrow=2)
mean(test)
colMeans(test)
mean(test2)
colMeans(test2)
```

Which command generates an error? Why? How is `mean()` different to `colMeans()`? Is the `length()`, when applied to a data frame or matrix, equal to the number of rows?

### Extracting rows and columns
After we have loaded/imported our data we want to be able to work with it. Typical operations are to extract specific columns, rows, or values, and to add new columns and rows. Recall that **columns** run top to bottom, while **rows** run left to right.

In addition, columns will sometimes have names:
```{r, include=F}
data <- read.csv(file = "shrub-volume-data.csv")
```
```{r}
names(data)
```
To extract a column, we use the symbol `$` followed by the name of the column (in this case `site`):
```{r}
  data$site
```
If the columns do not have names you can specify the column you require by using `data[,n]`, where `n` is the number of column you require (beginning from the left). When no number precedes the `,n` this means _all rows_:
```{r}
data[,1]
```
To extract row `n`, and all the columns, we use `data[n,]`. Row 1 (all columns) is extracted as:
```{r}
  data[1,]
```

If we want one specific value, `data[n,m]` returns the value in row `n` and column `m`:

```{r}
  data[1,2]
```
Furthermore, `data[,n:m]` returns all the rows for columns `n` to `m`:

```{r}
  data[2:4,]
```

while `data[n:m,]` returns all the columns for rows `n` to `m`:
```{r}
  data[,2:3]
```

Note that in the two examples above `n` must be bigger than `m`. This method is called _slicing_.

This are simple commands to work with data, but as data scientists frequently want to work with data in this way, the `dplyr` package was created to handle data more efficiently.


## dplyr
We recommend using `dplyr` to handle data as the commands are concise, making the code easier to read and write. [Install and load](https://ahurford.github.io/quant-guide-all-courses/rintro.html#r-packages) the `dplyr` R package.

```{r, include=F, echo=F}
library(dplyr)
```

The `dplyr` package contains several different function that are very useful when manipulating and cleaning large datasets. We will be highlighting the use of the `select()`, `mutate()`, `length()`, `arrange()`, and `filter()`.

### Selecting a column
The command `select(data, length)` selects the `length` column of the dataframe `data`:

```{r, echo=T}
select(data, length)
```

This achieves the same result as the `data$length` command we learned in the previous section.

### Adding a new column

There are several different methods to add new columns to a data set in R, but when using the `dplyr` package, the function `mutate()` is appropriate. Below we create a new column called `site_double` that is equal to the `site` column times 2.

```{r}
data2 = mutate(data, site_double = site * 2)
```

We can see that `data2` has a new column called `site_double`.

### Sorting the data

The `arrange()` function orders the rows of a data frame by the values in a specific column. Below we arrange `data` by increasing order of the `width` column.:

```{r}
arrange(data, width)
```


### Filtering data

The `filter()` function is used to subset a data frame, retaining all rows that satisfy your conditions. For a row to be retained, `TRUE` must be produced for all conditions. To extract all rows of `data` with length greater than 7 cm:

```{r}
filter(data, height > 7)
```

We can also filter with two (or more) conditions. We use `&` (meaning _and_) if we require _both_ conditions to be satisfied, or `|` (meaning _or_) if we require _at least one_ condition is satisfied. Below, we filter the `data` so that `data3` contains only plants that have length greater than 8 cm and width greater than 5 cm.

```{r}
data3=filter(data, height >8 & width > 5)
```

However, no plants met both conditions so `data3` is empty.

We can filter to retain plants from Site 1 or Site 3:

```{r}
filter(data, site == 1 |site == 3)
```

In R,  `==` evaluates whether something is `TRUE` or `FALSE`, while `=` assigns the value on the right hand side to the variable on the left hand side. **TRY IT** In the `Console` try the following:


```{r}
# Assigns x the value 0
x = 0
# Evaluates TRUE if x = 0 (TRUE since we assigned x=0 on the above line of code)
x == 0
# Evalues TRUE if x = 1 (FALSE since we assigned x=0 on the first line of code)
x == 1
```

Therefore, when we write `site == 1`, `TRUE` is returned when the value of `site` is `1`, and these rows are retained in the output. If we wrote `site = 1`, then `R` would try to assign the variable `site` with the value `1`, which would return an error message since the second argument of the `filter()` function requires a logical statement that evaluates as `TRUE` or `FALSE` and `site = 1` assigns a value; it is not a logical statement.

### Removing NA entries in data

 Often experimental data is not as neat or pretty as example data. There is often missing data, or empty or `NA` entries which may cause issues in graphing or data analysis. `R` does not allow empty values, so empty values in `.csv` or `.xlsx` files are often converted to `NA` when these data are uploaded into `R`. Luckily, again the `filter` function has a solution to this! We can filter the data to remove rows with `NULL` or `NA` entries.

```{r}
#filter the data to remove rows with null values in the height column
filter(data, !is.na(width))
```

Observe that row 12 of `data` has now been removed because it contained `NA` in the `width` column. To understand this command, note that `is.na(x)` returns `TRUE` if `x = NA`. The `!` means 'not' and reverses the `TRUE` and `FALSE` outputs. We illustrate `!` below:

```{r}
# 1 equal to 2 (FALSE)
1 == 2

# 1 is not equal to 2 (TRUE)
1 != 2
```


Recall, that `filter()` retains the rows that evaluate as `TRUE`: we wish to remove the rows with `NA`, so we need to write a command that evaluates these to `FALSE`. Therefore, we use `!is.na()`.

### Grouping
The function`group_by()` can be used with `summarize()` and `tally()`. Recall that our data consists of different sites:

```{r}
head(data)
```


The code below calculates the mean `length` for each `site` (because we group `data` by `site`):

```{r}
by_site <- group_by(data,site)
data2 <- summarize(by_site,length=mean(length))
print(data2)
```

Applying the function `group_by()` doesn't make the data noticeably different aside from noting the grouping:

```{r}
by_site <- group_by(data,site)
head(print(by_site))
```

The first argument,  `x`, of `summarize(x,y)` is the grouped data, and the second argument, `y`, has the structure `name of the new column` = `function to be applied to(column of grouped data)`. Recalling that the function [`min()`](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#useful-functions) calculates the minimum of its argument, we might also calculate the minimum width for each of the sites:

```{r}
summarize(by_site, min.width = min(width))
```

The function `tally()` counts the number of records in each group.
```{r}
tally(by_site)
```

Equivalently, we could apply the function `length()` using `summarize()`:
```{r}
summarize(by_site,n=length(site))
```

### Joining

A good description of joining data sets using `inner_join()` is provided [here](https://stat545.com/join-cheatsheet.html).

### Pipes

The pipe function, `%>%`,  allows for compact coding when the output on the lefthand size of `%>%` is the input for the calculations performed on the righthand size of `%>%`. Consider:

```{r}
data2 <- data %>%
         filter(!is.na(width)) %>%
         mutate(height_cm=height*100) %>%
         select(site, experiment, height_cm)
head(data2)
```

We end up with a new dataframe, `data2`, that is `data` with three operations sequentially performed. The advantage of using `%>%` is the compactness and clarity of the code to achieve this result. Let's understand these commands further.

- `data2` is the name of the output when we are finished all our operations;
- `data` is the data that we start with;
- `filter(!is.na(width))` removes all rows with `NA` in the `width` column of `data`. In the previous sections, when we demonstarted `filter(x,y)`, the first argument, `x`, was a dataframe, and the second argument was a logical statement, i.e., `!is.na(width)`. Above, when `%>%` is used only the logical statement is provided and the data that the function applies to is supplied on the left hand side of `%>%`.
- `mutate(height_cm=height*100)` converts height in meters to centimeters, and this operation is applied to all the data that passed through the pipe from the previous operation where `NA`s in the `width` column were removed.
- `select(site, experiment, height_cm)` is used to output only columns of interest, and again, the data argument to the `select()` function is supplied by the lefthand size of the pipe.

Note that the pipes will not work if the output of one operation is of the wrong type to be the input to the next operation.

To understand the value of pipes, consider the equivalent code written without using pipes:
```{r}
data2 <- filter(data,!is.na(width))
data3 <- mutate(data2,height_cm=height*100)
data4 <- select(data3, site, experiment, height_cm)
head(data4)
```

Without using pipes, we create variables `data2` and `data3` that only serve as intermediaries, that we may never use again in our subsequent lines of code. Creating many variables can bog down both the computer memory and your own memory, when you are trying to remember the variable name you gave to a particular output when writing subsequent commands.


## Reshaping Data

Data frequently come in formats that are unsuitable for analysis. You would be required to do modifications on the data in some way. The process of cleaning or modifying the data to make it ready for analysis is known as `tidying`, or `munging`. The goal is generally to get the data into a `tidy` format. You often get data where items that should be in the rows are in the columns or vice versa yet it's not in a transpose form. Some plotting functions in R have a special way the data should be formatted before it can produce the required plot. In these circumstances, you would have to reshape the data into a format that is easier to work with. This section considers the various ways we can format data into a wide or long format. Depending on your goal, you may want the data frame to be in one of these specific formats. We hereby going to learn how to reshape data into one of these formats. 

### Reshaping data from wide to long format

1. We use the **stack method** in base R to transform data available in the form of separate columns within a data frame or a list into a single column. The stack method uses the two functions `cbind()` and `stack()` produces a result in the form of a data frame with two columns. We will demonstrate how this method is done by creating a sample data frame as follows and following through the lines of code below.

```{r}
# creating a data frame
data <- data.frame(name=c('Josh', 'Josh', 'Jake',
                          'Jake', 'Gorge', 'Gorge'),
                   semester=c(rep(LETTERS[1:2],3)),
                   quiz_BIOL1001=c(1, 3, 4, 8, 9, 1),
                   quiz_BIOL1002=c(2, 4, 6, 2, 7, 3))

data
```

We would now bind the first two columns of the data frame named `data` and stake the third and fourth columns and save it in a new name called `data_long`. 

```{r}
data_long <- cbind(data[1:2], stack(data[3:4]))
```

You can query the the new data created in the console to see the content.

```{r}
data_long
```

2. Using **melt method** in `reshape2` package in R. As we already know, we will have to install and load the `reshape2` package before it can be used in R if not already done. 

```{r}
#install.packages("reshape2")
library(reshape2)
```


Now we will see how the melt method in this package can be used to stack data frame columns together by going through an example. You can read more about the syntax for the `melt()` function by typing *?melt* in the console. We would use the data frame named `data` for this ecample.

```{r}
# binding the first two columns as it is and stacking the third and fourth columns
data_melt <- reshape2::melt(data, id.var = c('name', 'semester'),
                           variable.name = 'course_name')

print(data_melt)
```

Here is another example on how to use the `melt()` function in the `reshape2` package to reshape data.

```{r}
#create data frame
data2 <- data.frame(person=c('A', 'A', 'B', 'B', 'C', 'C'),
                   trial=c(1, 2, 1, 2, 1, 2),
                   outcome1=c(7, 6, 6, 5, 4, 4),
                   outcome2=c(4, 4, 5, 5, 3, 2))

data2
```

Reshape `data2` from the wide format to a long format by staking the third and fourth columns.

```{r}
#melt columns of data frame
data2_long <- melt(data2, id.var = c('person', 'trial'), variable.name = 'outcome')

data2_long
```

3. Using the **tidyr package**. The `tidyr` package make use of the following functions:

`pivot_longer()`: Reshapes a data frame from wide to long format.

`pivot_wider()`: Reshapes a data frame from long to wide format.

We will study how to use `pivot_longer()` function in the next example and study how to use `pivot_wider()` function in the section that follows. 

To use these functions, we would be required to run `install.packages("tidyr")` to get the package installed if you have not done that previously. Also remember to load the package. I would also use  the pipe operator, `%>%` in the code. `%>%` is not required to use `tidyr`, but it does make things easier. To use the pipe operator, you must also install `dplyr` package.

```{r}
# I have "commented out" these two lines of code with the # sign. Just take away the # and run the lines. 
#install.packages("tidyr") 
#install.packages("dplyr")

library(tidyr)
library(dplyr)

#pivot the data frame into a long format. Here we use the pipe function. 
data %>% pivot_longer(cols=c('quiz_BIOL1001', 'quiz_BIOL1002'),
                    names_to='quiz',
                    values_to='marks')
```

Another example with `data2` data frame using the `pivot_longer()` function in the `tidyr` package. 

```{r}
#pivot the data frame into a long format.
data2 %>% pivot_longer(cols=c('outcome1', 'outcome2'),
                    names_to='outcomes',
                    values_to='values')
```

4. Another function of `tidyr` that can be used to transform the data from wide to long is the `gather()` function. We again show how to use this function.

```{r}
df_long <- data %>% gather(quiz, marks, quiz_BIOL1001:quiz_BIOL1002)
df_long
```

### Reshaping data from long to wide format

1. We will be using the **tidyr package** to reshape data from long to wide. Let us study how to use `pivot_wider()` function in the next example to pivot data frame into a wide format. We will be using the example data frame created in the previous examples. 

```{r}
#pivot the data frame into a wide format
df_wide <- df_long %>% pivot_wider(names_from = quiz, values_from = marks)
df_wide
```

2. Another function of `tidyr` that can be used to transform the data from long back to wide is the `spread()` function. We show how to use this function.

```{r}
df_wide2 <- df_long %>% spread(quiz, marks)
df_wide2
```

We show another example of using the `spread()` function to transform back `data2` from long to wide. 

```{r}
data2_long %>% spread(outcome, value)
```

There are many other functions to reshape data in R but these are the few handy ones that you can always stick to in your data analysis and plotting. 



