# Handling data

Base R refers to functions that come with an installation of `R`, not those that are in packages that are must be installed. We will learn how to handle data both using the base R functions, and using the package `dplyr`.

## Base R

### Useful functions

-   `max()` outputs the maximum value in a vector (recall, a vector is a list of numbers, e.g., `c(0,2,5)`), matrix, or data frame;

-   `min()` outputs the minimum value in a vector, matrix, or data frame;

-   `length()` outputs the total number of elements, in a vector, matrix, or data frame;

-   `sum()` adds or takes the sum of all values in a vector, matrix, or data frame;

-   `mean()` calculates the mean of all values in a vector or matrix;

-   `colMeans()` calculates the mean of a column in a matrix or data frame.

- `head()` prints the first 6 rows of the argument to your console. This can be helpful if you want to check your data without having to sort through a long output.

- `tail()` similar to `head()` but prints the last 6 rows of the argument.

**TRY IT!** In the `Console` do some tests on these functions by creating a test matrix and a test data frame:

```{r}
test = data.frame(col1 = c(0,2), col2 = c(5,6))
test2 = matrix(c(2,3),c(5,1),nrow=2)
mean(test)
colMeans(test)
mean(test2)
colMeans(test2)
```

Which command generates an error? Why? How is `mean()` different to `colMeans()`? Is the `length()`, when applied to a data frame or matrix, equal to the number of rows?

### Extracting rows and columns
After we have loaded/imported our data we want to be able to work with it. Typical operations are to extract specific columns, rows, or values, and to add new columns and rows. Recall that **columns** run top to bottom, while **rows** run left to right.

In addition, columns will sometimes have names:
```{r, include=F}
data <- read.csv(file = "shrub-volume-data.csv")
```
```{r}
names(data)
```
To extract a column, we use the symbol `$` followed by the name of the column (in this case `site`):
```{r}
  data$site
```
If the columns do not have names you can specify the column you require by using `data[,n]`, where `n` is the number of column you require (beginning from the left). When no number precedes the `,n` this means _all rows_:
```{r}
data[,1]
```
To extract row `n`, and all the columns, we use `data[n,]`. Row 1 (all columns) is extracted as:
```{r}
  data[1,]
```

If we want one specific value, `data[n,m]` returns the value in row `n` and column `m`:

```{r}
  data[1,2]
```
Furthermore, `data[,n:m]` returns all the rows for columns `n` to `m`:

```{r}
  data[2:4,]
```

while `data[n:m,]` returns all the columns for rows `n` to `m`:
```{r}
  data[,2:3]
```

Note that in the two examples above `n` must be bigger than `m`. This method is called _slicing_.

This are simple commands to work with data, but as data scientists frequently want to work with data in this way, the `dplyr` package was created to handle data more efficiently.


## dplyr
We recommend using `dplyr` to handle data as the commands are concise, making the code easier to read and write. [Install and load](https://ahurford.github.io/quant-guide-all-courses/rintro.html#r-packages) the <u>[dplyr](rdocumentation.org/packages/dplyr/versions/0.7.8)</u> R package.

```{r, include=F, echo=F}
library(dplyr)
```

The [dplyr](rdocumentation.org/packages/dplyr/versions/0.7.8) package contains several different function that are very useful when manipulating and cleaning large datasets. We will be highlighting the use of the `select()`, `mutate()`, `length()`, `arrange()`, and `filter()`.

### Selecting a column
The command `select(data, length)` selects the `length` column of the dataframe `data`:

```{r, echo=T}
select(data, length)
```

This achieves the same result as the `data$length` command we learned in the previous section.

### Adding a new column

There are several different methods to add new columns to a data set in R, but when using the `dplyr` package, the function `mutate()` is appropriate. Below we create a new column called `site_double` that is equal to the `site` column times 2.

```{r}
data2 = mutate(data, site_double = site * 2)
```

We can see that `data2` has a new column called `site_double`.

### Sorting the data

The `arrange()` function orders the rows of a data frame by the values in a specific column. Below we arrange `data` by increasing order of the `width` column.:

```{r}
arrange(data, width)
```


### Filtering data

The `filter()` function is used to subset a data frame, retaining all rows that satisfy your conditions. For a row to be retained, `TRUE` must be produced for all conditions. To extract all rows of `data` with length greater than 7 cm:

```{r}
filter(data, height > 7)
```

We can also filter with two (or more) conditions. We use `&` (meaning _and_) if we require _both_ conditions to be satisfied, or `|` (meaning _or_) if we require _at least one_ condition is satisfied. Below, we filter the `data` so that `data3` contains only plants that have length greater than 8 cm and width greater than 5 cm.

```{r}
data3=filter(data, height >8 & width > 5)
```

However, no plants met both conditions so `data3` is empty.

We can filter to retain plants from Site 1 or Site 3:

```{r}
filter(data, site == 1 |site == 3)
```

In R,  `==` evaluates whether something is `TRUE` or `FALSE`, while `=` assigns the value on the right hand side to the variable on the left hand side. **TRY IT** In the `Console` try the following:


```{r}
# Assigns x the value 0
x = 0
# Evaluates TRUE if x = 0 (TRUE since we assigned x=0 on the above line of code)
x == 0
# Evalues TRUE if x = 1 (FALSE since we assigned x=0 on the first line of code)
x == 1
```

Therefore, when we write `site == 1`, `TRUE` is returned when the value of `site` is `1`, and these rows are retained in the output. If we wrote `site = 1`, then `R` would try to assign the variable `site` with the value `1`, which would return an error message since the second argument of the `filter()` function requires a logical statement that evaluates as `TRUE` or `FALSE` and `site = 1` assigns a value; it is not a logical statement.

### Removing NA entries in data

 Often experimental data is not as neat or pretty as example data. There is often missing data, or empty or `NA` entries which may cause issues in graphing or data analysis. `R` does not allow empty values, so empty values in `.csv` or `.xlsx` files are often converted to `NA` when these data are uploaded into `R`. Luckily, again the `filter` function has a solution to this! We can filter the data to remove rows with `NULL` or `NA` entries.

```{r}
#filter the data to remove rows with null values in the height column
filter(data, !is.na(width))
```

Observe that row 12 of `data` has now been removed because it contained `NA` in the `width` column. To understand this command, note that `is.na(x)` returns `TRUE` if `x = NA`. The `!` means 'not' and reverses the `TRUE` and `FALSE` outputs. We illustrate `!` below:

```{r}
# 1 equal to 2 (FALSE)
1 == 2

# 1 is not equal to 2 (TRUE)
1 != 2
```


Recall, that `filter()` retains the rows that evaluate as `TRUE`: we wish to remove the rows with `NA`, so we need to write a command that evaluates these to `FALSE`. Therefore, we use `!is.na()`.

### Grouping
The function`group_by()` can be used with `summarize()` and `tally()`. Recall that our data consists of different sites:

```{r}
head(data)
```


The code below calculates the mean `length` for each `site` (because we group `data` by `site`):

```{r}
by_site <- group_by(data,site)
data2 <- summarize(by_site,length=mean(length))
print(data2)
```

Applying the function `group_by()` doesn't make the data noticeably different aside from noting the grouping:

```{r}
by_site <- group_by(data,site)
head(print(by_site))
```

The first argument,  `x`, of `summarize(x,y)` is the grouped data, and the second argument, `y`, has the structure `name of the new column` = `function to be applied to(column of grouped data)`. Recalling that the function [`min()`](https://ahurford.github.io/quant-guide-all-courses/handling-data.html#useful-functions) calculates the minimum of its argument, we might also calculate the minimum width for each of the sites:

```{r}
summarize(by_site, min.width = min(width))
```

The function `tally()` counts the number of records in each group.
```{r}
tally(by_site)
```

Equivalently, we could apply the function `length()` using `summarize()`:
```{r}
summarize(by_site,n=length(site))
```


### Pipes

The pipe function, `%>%`,  allows for compact coding when the output on the lefthand size of `%>%` is the input for the calculations performed on the righthand size of `%>%`. Consider:

```{r}
data2 <- data %>%
         filter(!is.na(width)) %>%
         mutate(height_cm=height*100) %>%
         select(site, experiment, height_cm)
head(data2)
```

We end up with a new dataframe, `data2`, that is `data` with three operations sequentially performed. The advantage of using `%>%` is the compactness and clarity of the code to achieve this result. Let's understand these commands further.

- `data2` is the name of the output when we are finished all our operations;
- `data` is the data that we start with;
- `filter(!is.na(width))` removes all rows with `NA` in the `width` column of `data`. In the previous sections, when we demonstarted `filter(x,y)`, the first argument, `x`, was a dataframe, and the second argument was a logical statement, i.e., `!is.na(width)`. Above, when `%>%` is used only the logical statement is provided and the data that the function applies to is supplied on the left hand side of `%>%`.
- `mutate(height_cm=height*100)` converts height in meters to centimeters, and this operation is applied to all the data that passed through the pipe from the previous operation where `NA`s in the `width` column were removed.
- `select(site, experiment, height_cm)` is used to output only columns of interest, and again, the data argument to the `select()` function is supplied by the lefthand size of the pipe.

Note that the pipes will not work if the output of one operation is of the wrong type to be the input to the next operation.

To understand the value of pipes, consider the equivalent code written without using pipes:
```{r}
data2 <- filter(data,!is.na(width))
data3 <- mutate(data2,height_cm=height*100)
data4 <- select(data3, site, experiment, height_cm)
head(data4)
```

Without using pipes, we create variables `data2` and `data3` that only serve as intermediaries, that we may never use again in our subsequent lines of code. Creating many variables can bog down both the computer memory and your own memory, when you are trying to remember the variable name you gave to a particular output when writing subsequent commands.
