# **PART II: ADVANCED** {-}

# Writing functions
## Introduction

Writing your own functions allows you to automate common tasks rather than copy-and-pasting. Functions reduce duplication by identifying repeated patterns of code and extract them out into independent pieces that can be easily reused and updated. This is beneficial for several reasons.

If you need to update your code, you only need to do so in one place.

You're less likely to make inadvertent mistakes when writing functions compared to when you copy and paste. For example, when updating a variable or name in one section of a copy-and-pasted script, you will need to remember to change it in all the others.

It will tidy and shorten your R scripts.

## Writing your own functions

If you find yourself having to repeat a task multiple times, it can be helpful to create your own custom function. The syntax of a function is given below:

```
function_name <- function(argument1, argument2, ...) {
    statement or code that completes your task # function body
    return(value)
}
```

function_name: This is the name you give your function. It should be relatively short but describe what the function does. Try avoid using function names that are already used elsewhere in R (e.g., sum). 

argument1, argument2, …: The arguments of the function are any R objects (numeric, strings, characters, data.frames, matrices, other functions). You can have any number of arguments

function body: The code between the '{}' is the function body and uses the arguments to complete the specified tasks. Every time the function is called this code is run.

return value: The last line of code is the object to be returned. 

Using this format, we can write a simple function to calculate the squared value of object x.

```{r}
square_num <- function(x) {
  square <- x^2
  return(square)
}
```

Now, we can use 'square_num' as we would any other function. We type out the name of the function, and inside the parentheses we provide a numeric value for x:

```{r}  
square_num(5)
```

We can write a function with multiple arguments as well.

```{r}
add_num <- function(a, b, c) {
  a * b + c
}
add_num(3,10,8)
```

We can also use functions to calculate a range of values. For example, we might be given a set of temperature values in degrees Celsius that we need to change to Fahrenheit. 

```{r}
temp_C <- c(0, 5, 10, 15, 20, 25, 30)

celsius_to_fahrenheit <- function(temp_C) {
  temp_F <- (temp_C * 9/5) + 32
  return(temp_F)
}

celsius_to_fahrenheit(temp_C)
```

## Returning Multiple Outputs

The examples above all return a single item. What if I want to return multiple items from a function? Let's start with a simple example that is similar to what we have already covered. The following function, which we have called `dice`, simulates rolling a six-sided dice by randomly sampling numbers 1 to 6 `n` times with replacement.

```{r}
dice <- function(n){
    roll <- sample(1:6, n, replace = T)
    return(roll)
}

```

Choosing to "roll" the dice 10 times gives us the following output.

```{r}
dice(10)
```

Now instead of the dice function returning the result of each individual roll, we now want a) a table with the total number of times each result occurred, and b) to plot the results. We can add a new line to the function for each argument. First we construct a table called 'total' that provides the summed results, then we plot these totals.

```{r}
dice <- function(n){
    toss <- sample(1:6, n, rep = T)
    total <- table(toss)
    barplot(total, xlab = "Dice Number", ylab = "Total Rolls", main = "")
    return(total)
}

dice(10)
```

# Loops and repeating commands

## Introduction
Loops are used in programming to repeat a specific block of code. Similar to writing functions, this approach will reduce repetition and keep your code organized. 

## for loops 
A for loop is used to iterate over a vector in R programming. Iteration helps apply the same task to multiple inputs, in other words, for loops perform the same action for each item in a list. 

The syntax of loops is relatively simple - for each value listed within the `for()` component, the operation stated within the `{}` is performed.

```
for(value in sequence)
{
statement
}
```

In the for loop below, we have specified a vector containing the numbers 1 to 5. The statement within the `{}` will use the `print` function to return the value of `i` with each loop or iteration. In this example, the loop iterates 5 times as the vector within `for()` contains 5 elements. 

```{r}
for (i in 1:5) {
  print(i)
}

```

Instead of creating the vectors within `for()`, we can also use existing vectors. We first assign the vector we are looping over before calling the loop. In the example below, the loop will return the squared value for each term within x.

```{r}
x <- 1:5
for (i in x) {
  print(i^2)
}

```

Often we will want to save the output from our loop. The first option is to make a blank vector or data frame and save the results to it. 

In the example below, we first create an empty vector `x`. The for loop first calculates the square of the numbers 1 to 10 and then saves them to vector `x`.

```{r}
x <- vector() 

for (i in 1:10) {
  y <- i^2
  x <- c(x, y)
}

x
```

If you know the dimensions of your data, you can also make a blank vector or dataframe of a specified size. This can help increase processing speeds for more complicated loops. For example, if we had a loop with 10 elements, we could store the results of each operation in a vector with a length of 10

Here is code that will store both the square and the square root of the numbers 1 to 10 in two columns of a new dataframe called x2

```{r}
x2 <- data.frame(col1 = vector(length = 10), col2 = vector(length = 10)) 

for (i in 1:10) {
  col1 <- i^2 
  col2 <- sqrt(i) 
  x2[i, 1] <- col1 
  x2[i, 2] <- col2 
}

x2
```

Now let's say we are given a dataset that includes three experiments across four different sites.

```{r, include=F}
data <- read.csv(file = "shrub-volume-data.csv")
```

```{r}
data
```

We are then asked to calculate the correlation between shrub length and height for experiment 1 and 2. We first create a vector that specifies these experiment numbers and then create an empty vector that will be used to store the correlation results.

```{r}
exp_num <- c(1, 2)

cor_output <- c()
```

Within our `for()` loop we will need to include several steps. First we subset our data using the values previously defined in `exp_num` above and store as a variable, which we call `exp_subset`. We then calculate the correlation between length and height from `exp_subet`. Next we create a dataframe that will include both the experiment number as well as the correlation results and finally, we use the function `rbind` to bind the rows from `df_cor` with our empty vector. 

```{r}
for(i in exp_num){
  exp_subset <- subset(data, experiment == exp_num[i])
  cor_val <- cor(exp_subset$length, exp_subset$height) 
  df_cor <- data.frame(experiment = i,
                       correlation = cor_val)
  cor_output <- rbind(cor_output, df_cor) 
}
cor_output
```

**TRY IT** How out would rewrite the above code if you wanted to instead subset for sites 2 and 4?

## Loop alternatives

Loops, specifically for-loops, are essential to programming in general. However, in R, you should avoid them when there are built-in functions that already exist. These built-in functions are often more efficient ways of doing things rather than writing a loop yourself.

One option is to use the `apply()` family of functions. These functions manipulate data from matrices, arrays, lists and dataframes in a repetitive way while avoiding explicit use of loops. Here is an example of some of the more common functions within the `apply()` family:

* `apply()` -	Applies a function over over rows or columns of an array or matrix.
* `lapply()` - Apply a function over a vector or list and returns a list.
* `sapply()` - Similar to `lapply()` but can simplify the result to a vector or matrix.

# ggplot

## Introduction

ggplot2 is an R package for producing data graphics. Unlike the base plot function we learned earlier, ggplot2 creates graphs using a step-by-step process that combines independent layers. This makes ggplot2 a very flexible and powerful tool for producing statistical graphics.

Every plot made using ggplot2 has three key components:

1. Data: a data frame

2 Aesthetic mappings: creates the link between data variables (x and y) and visual properties (colour, size or shape of points, etc)

3. Geometry (geom) functions: at least one layer that describes the type of graphic

We link these components together to create our plot: plot = data + aesthetics + geometry

## Making scatter plots with ggplot2

We first load the `ggplot2` package and import our dataset. 

```{r, include=F}
ShrubData <- read.csv(file = "shrub-volume-data.csv")
ShrubData$experiment <- as.factor(ShrubData$experiment)
```

```{r}
library(ggplot2)
head(ShrubData)
```

We begin to build our graphic one step at a time, first using the function `ggplot` to specify our data frame and set our x and y variables within the aesthetic `aes()`. When we run this code, it produces a blank figure with our x and y axis set because we haven't specified the type of graphic we want.

```{r}
ggplot(data = ShrubData, aes(x = length, y = height))
```

To add the data, we must choose a geom function, for example `geom_point()`.

```{r, warning = FALSE}
ggplot(data = ShrubData, aes(x = length, y = height)) + 
  geom_point()

```

If we want to change the colour or shape of the points to represent the specific site, we can add that to the aesthetic within `geom_point`.

```{r, warning = FALSE}
ggplot(data = ShrubData, aes(x = length, y = height)) + 
  geom_point(aes(colour = experiment, shape = experiment))
```

**TRY IT!** Set `colour = blue` both inside `aes()` and outside it. How does your graph change?

Hint: Specifying colour within the aesthetic allows you to set variable colour according to a class. If you want to set a fixed variable colour, you do this outside the aesthetic. 

## Customising graphs

To change any aspect of your graph (e.g., colour, shape, line type, font, etc.), you can build off the existing plot. We do this by adding an additional layer for each element we want to change. 

For example, if we want to change the colour or shape of the points, we can run the following code:

```{r}
ggplot(data = ShrubData, aes(x = length, y = height)) + 
  geom_point(aes(colour = experiment, shape = experiment)) +
  scale_color_manual(values=c("blue", "orange", "purple")) + 
  scale_shape_manual(values=c(21, 23, 22))
```

We can also change our axis title using the `xlab()` or `ylab()` functions.
```{r}
ggplot(data = ShrubData, aes(x = length, y = height)) + 
  geom_point(aes(colour = experiment, shape = experiment)) + 
  xlab("Shrub Length (cm)") + 
  ylab("Shrub Height (cm)") 
```

Similarly, we can change the limits of the axes using `xlim()` or `ylim()`.
```{r}
ggplot(data = ShrubData, aes(x = length, y = height)) + 
  geom_point(aes(colour = experiment, shape = experiment)) + 
  xlim(0, 5) + 
  ylim(0, 10) 
```

All aspects of a ggplot graph are customizable, making it a very flexible tool. More information can be found here:https://ggplot2.tidyverse.org/articles/faq-customising.html 

## Adding lines of best fit

We can continue to build layers to our graphic by adding additional geoms. For example, if we want to add a line of best fit we use `geom_smooth`, specifying `method = "lm"` for a linear model. We also add `colour = experiment` within the `aes()` to ensure we get one line for each experiment. 

```{r, message = FALSE, warning = FALSE}
ggplot(data = ShrubData, aes(x = length, y = height)) + 
  geom_point(aes(colour = experiment, shape = experiment)) + 
  geom_smooth(aes(colour= experiment), 
              method= "lm", formula = y ~ x, se = FALSE)
```

**TRY IT!** What happens if we don't include the `aes()` within `geom_smooth()`?

## Facet wrapping
Facet wrapping allows you to view each category on its own. 

For example, if you wanted to make a separate graph for each experiment, you can add `facet_wrap(~experiment)`. Here the tilde (~) indicates which category you want to separate your data by.

```{r, message = FALSE, warning = FALSE}
ggplot(data = ShrubData, aes(x = length, y = height)) + 
  geom_point(aes(colour = experiment, shape = experiment)) + 
  geom_smooth(aes(colour= experiment), 
              method= "lm", formula = y ~ x, se = FALSE) + 
  facet_wrap(~experiment)
```

## Other types of plots

We can visualize the data using other geom functions such as `geom_line` or `geom_hist`.

``` {r}
ggplot(data = ShrubData, aes(x = length, y = height)) + 
  geom_line(aes(colour = experiment, linetype = experiment))
```

``` {r}
ggplot(data = ShrubData, aes(x = experiment, y = height)) + 
  geom_boxplot(aes(fill= experiment)) 
```

## Further reading
* The ggplot2 package has a website with many help articles including other geom types and how to customize your graphs (fonts, colours, shapes, etc.) https://ggplot2.tidyverse.org/index.html

# Making maps
```{r, include = FALSE}
### Install Packages ----
## An R package to hold and facilitate interaction with Natural Earth map data.
if (!require(rnaturalearth)) install.packages("rnaturalearth")

## Data to support much of the package functionality are stored in two additional data packages
if (!require(rnaturalearthdata)) devtools::install_github("ropensci/rnaturalearthdata")
if (!require(rnaturalearthhires)) devtools::install_github("ropensci/rnaturalearthhires")

## Load Library
libs <- c('ggplot2', "sf", "rnaturalearth", "rnaturalearthdata", "rnaturalearthhires", "ggspatial", "raster", "dplyr", "cowplot")
lapply(libs, require, character.only = TRUE)

```
## Introduction

Maps are a great tool for presenting data in a simplified and visually appealing way. Typically, maps are created using GIS (Geographical Information System) software such as ArcGIS or QGIS, however in recent years R has developed many packages that can process and present spatial data. Drawing maps in R provides reproducible and flexible code that allows us to customize each map for a specific audience or purpose.

In GIS, there are a few key spatial elements that we must understand to build a map:

* Points: Specific locations with a single pair of coordinates (e.g., latitude and longitude) such as a city or town.
* Lines: Linear shapes that are made of connected point data such as roads or rivers. Lines use an ordered set of coordinates. 
* Polygons: Closed shapes made of connected lines such as country or provincial/territorial borders. Polygons also have an ordered set of coordinates but the start and end point should have the same coordinates.

Each of these elements can be added as their own layer when building map.

## Creating maps using `ggplot2` and `sf` 

We can now build off of our ggplot knowledge to build maps in R. We will also need to install the package `sf`. `sf` provides a standardized way to work spatial vector data (i.e., points, lines, and polygons) by encoding the data as "simple features". Simple features also keep any associated data as a data.frame, which allows us to change the aesthetic of each feature when working with ggplot.

To build our example map, we will be using data from the `rnaturalearth` package as well as the associated data packages `rnaturalearthdata` and `rnaturalearthhires`. 

Once we have loaded these packages we can use the function `ne_states()` to select a map of Canada, as well as all of the provincial and territorial boundaries. Note: we must set `returnclass = "sf"` to make sure the output is a "simple feature".

```{r}
canada_map <- ne_states(country = "canada", returnclass = "sf")

class(canada_map) 

```

Using the function `class` tells us that the object `canada_map` is both a `sf` and a `data.frame`.

We can now plot our base map using `ggplot()` and adding a layer called `geom_sf()` where we specify the data we will be using.

```{r}
ggplot() + 
  geom_sf(data = canada_map) 
```

## Changing map colours

We can now start to customize our map, first by changing the colours. To change the colour of the borders we specify this using `colour = `, and to change the colour of the interior, we can specify this using `fill - `. Note: we set these values outside of the aesthetic (`aes()`) because we want to apply them to the entire object of `canada_map`. 

```{r}
ggplot() + 
  geom_sf(data = canada_map, colour = "black", fill = "white") 

```

We can also set the fill inside of `aes()` if we want it to be based on values found within the dataframe. For example, we can set the fill by the name of each Canadian province or territory.

```{r}
ggplot() + 
  geom_sf(data = canada_map, colour = "black", aes(fill = name))
```

If we instead wanted to highlight one province, for example Newfoundland and Labrador, we again set this outside of `aes()`. We also will use an `ifelse()` statement. In the code below, this statement can be interpreted as "*if* the name from our dataframe matches Newfoundland and Labrador return the colour blue, *else* return the colour white.

```{r}
ggplot() + 
  geom_sf(data = canada_map, colour = "black", 
          fill = ifelse(canada_map$name == "Newfoundland and Labrador", 'blue', 'white')) 
```

Lastly, we can change the background fill to the map using `theme` elements. The layer `theme_bw()` will first set a black and white theme for the map, and can then alter the fill by setting `panel.background`. 

```{r}
ggplot() + 
  geom_sf(data = canada_map, colour = "black", fill = "white") +
  theme_bw() +
  theme(panel.background = element_rect(fill = "aliceblue"))
```
## Projections

The Earth exists in three-dimensions but, other than globes, most representations of the earth, such as maps, are two dimensional. A map projection is the process of representing Earth's surface in two dimensions using a set of mathematical transformations. However, the Earth's surface cannot be flattened without distorting at least one geometrical property, such as area, shape, and distance. You cannot accurately represent all of these properties simultaneously therefore, it is important to select a projection that has the most suitable properties for that the mapping situation. 

Projections in R can be specified using Well Known Identifier (WKID) codes or PROJ.4 strings. WKID codes are numeric identifiers of projections and are defined using either the ESPG or ESRI system. More information can be found following the links in the Additional Reading section.

We will use instead use a PROJ.4 string, which specifies the complete set of projection parameters including the ellipse, datum, and units that define a particular coordinate reference system (CRS). The PROJ.4 string for any given CRS can be found here: https://spatialreference.org.

For example, maps of Canada often use the Lambert Conformal Conic projection, which helps preserve both relative shape and direction. We will first copy the entire string from the website and save it as an object in R, which we will call `can.lcc`. We can then specify that we want to use this projection in for our map by adding a layer called `coord_sf()` and setting our CRS as `can.lcc`.

```{r}
can.lcc <- "+proj=lcc +lat_1=49 +lat_2=77 +lon_0=-91.52 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"

ggplot() + 
  geom_sf(data = canada_map, colour = "black", fill = "white") + 
  coord_sf(crs = can.lcc)
```

Compared to any of the maps prior to this, you will notice that the latitude and longitude grid have become conical and that the shape and size of the provinces and territories have been altered. 

## Map Extent

We may want to map a specific area, such as the province of Newfoundland and Labrador rather than the entirety of Canada. We can accomplish this by defining by a set of bounding coordinates that specify the area of the map, also called the "extent". We can specify the coordinate extent by setting the `xlim()` and `ylim()` values within the `geom_sf()` layer.

```{r}
ggplot() +
  geom_sf(data = canada_map, colour = "black", fill = "white") +
  coord_sf(xlim = c(-67.5, -53), ylim = c(47, 60.5))
```

Alternatively, we can subset the Canada map by selecting that row of data from our data.frame and saving it as a new object. We can then map just the province without any additional borders.

```{r}
map_nl <- canada_map[canada_map$name == "Newfoundland and Labrador",] 

ggplot() +
  geom_sf(data = map_nl, colour = "black", fill = "white") 
```

Lastly, we can specify the values we want in our x or y axes by setting breaks in the data. For example, we can reduce the number of longitude values included by setting `scale_x_continuous`.

```{r}
ggplot() +
  geom_sf(data = map_nl, colour = "black", fill = "white") +
  scale_x_continuous(breaks=seq(-68, -52, 3)) 
```

## Mapping points

We can add layers to our base map using additional `geom` functions in the same way that we would for any ggplot item. For points, we would use the function `geom_point()` and specify the coordinates as the x and y values within `aes()`. For this example, we will be mapping the populations of towns across Newfoundland and Labrador using the 2021 census report (Statistics Canada. Table 98-10-0002-01  Population and dwelling counts: Canada and census subdivisions (municipalities)).

```{r include = FALSE}
# Import Dataset
library(data.table)
nl.cities <- fread("2021_Can_Census_w_Locs.csv")
options(scipen=10000)
```

Once we have our data, we will use the same base map that we used prior, and add the `geom_point()` layer. Note that we have also specified the colour of the points to be based on the 2021 population of the town/city and that we have done so within the `aes()`.

```{r}
ggplot() +
  geom_sf(data = map_nl, colour = "black", fill = "white") +
  geom_point(data = nl.cities, aes(x = Longitude, y = Latitude, colour = Population_2021)) +
  scale_x_continuous(breaks=seq(-68, -52, 3)) 
```
This map doesn't provide a lot of information about the population sizes in NL since most towns have fewer than 25 000 people. To create a more meaningful representation of our data, we can change the scale and colour. 

One option is to add a layer using the function `scale_color_viridis_c()`, which provides a range of colour blind friendly palettes. We will just use the default option in this example.

We can also scale the data by specifying the transformation we would like to use; in the example below, we will use the `log10` transformation, which returns the base 10 logarithm of a number. Note that this transformation is for the colour scale only and it will not alter the values of our data. The legend will now provide a wider range of colours for values between 10 and 10 000.
Lastly, we will change the name of the legend within the `labs()` function.

```{r warning = FALSE}
ggplot() +
  geom_sf(data = map_nl, colour = "black", fill = "white") +
  geom_point(data = nl.cities, aes(x = Longitude, y = Latitude, colour = Population_2021)) + 
  scale_x_continuous(breaks=seq(-68, -52, 3)) +
  scale_color_viridis_c(trans = "log10") +
  labs(colour = "Population 2021") 
```

Other spatial elements such as lines or polygons can be added in the same way using the `geom_line()` or `geom_polygon()` functions.

## Scale bar and north arrow

Scale bars provide a visual representation of distance and feature size on the map. A scale bar is a line or bar divided into parts.

To add a scale bar or north arrow, we will need to load the R package `ggspatial`.
We use `annotation_scale()` to add a scale bar to the map. The location can be stated as `"tr"` for the top right corner, `"bl"` for bottom left corner, etc. The argument `width_hint` sets the size of the scale bar as a proportion of the map width. For example `width_hint = 0.5` will return a scale bar that is half the width of the map.

```{r message=FALSE}
ggplot() +
  geom_sf(data = map_nl, colour = "black", fill = "white") +
  scale_x_continuous(breaks=seq(-68, -52, 3)) +
  annotation_scale(location = "tr", width_hint = 0.5)

```

A north arrow indicates the orientation of the map inside the frame. Using the argument `annotation_north_arrow()`, we can specify the orientation using `which_north`. Options include "grid", where the north arrow always pointing up and "true", which always points to the north pole. We can also define the size of the north arrow using `height` and `width`, as well as the style and location.

```{r message=FALSE}
ggplot() +
  geom_sf(data = map_nl, colour = "black", fill = "white") +
  scale_x_continuous(breaks=seq(-68, -52, 3)) +
  annotation_scale(location = "tr", width_hint = 0.5) +
  annotation_north_arrow(location = "tl", which_north = "true", 
                         height = unit(1, "cm"), width = unit(1, "cm"),
                         style = north_arrow_fancy_orienteering)

```

## Finalizing the map

You may need to save the map for a report, presentation, or thesis. We will combine each of the previous steps into one map below, which we will assign as the object named `final`. These previous steps include:

* Plot a base map of Newfoundland and Labrador
* Add points representing the population size of towns/cities in Newfoundland
* Add a scale bar and north arrow
* Set the background colour to blue

In addition, we will move the legend inside the map margins by defining `legend.position` within the `theme` layer. 

```{r message = FALSE, warning = FALSE, fig.dim = c(7,7)}
final <- ggplot() +
  geom_sf(data = map_nl, colour = "black", fill = "white") +
  geom_point(data = nl.cities, aes(x = Longitude, y = Latitude, colour = Population_2021)) + 
  scale_x_continuous(breaks=seq(-68, -52, 3)) +
  scale_color_viridis_c(trans = "log10") +
  labs(colour = "Population 2021") +
  annotation_scale(location = "tr", width_hint = 0.5) +
  annotation_north_arrow(location = "tl", which_north = "true", 
                         height = unit(1, "cm"), width = unit(1, "cm"),
                         pad_x = unit(0.25, "cm"), pad_y = unit(0.25, "cm"),
                         style = north_arrow_fancy_orienteering) +
  theme_bw() +
  theme(panel.background = element_rect(fill = "aliceblue"),
        legend.position = c(0.85, 0.75), legend.background = element_rect(fill = "transparent"),
        legend.title = element_text(size = 10, face = "bold")) 
final
```

We may also want to create an inset map, which shows the location of the main map in the context of a larger area. For example, we can use the larger Canadian map we made previously, with Newfoundland and Labrador highlighted. We will assign this map as the object `inset`. Reminder, the previous steps included:

* Using an `ifelse` statement to colour Newfoundland and Labrador blue and the remainder of Canada white
* Reprojected the map using the Lambert Conformal Conical projection
* Set the map theme and changed the background colour

Now we will also add a scale bar to our inset map using `annotation_scale()` and remove the white background by setting `plot.background = element_blank()`. We will also change the map theme to `theme_void()` to remove the latitude and longitude and help simplify the map inset.

```{r}
inset <- ggplot() + 
  geom_sf(data = canada_map, colour = "black", 
          fill = ifelse(canada_map$name == "Newfoundland and Labrador", 'blue', 'white')) + 
  coord_sf(crs = "+proj=lcc +lat_1=49 +lat_2=77 +lon_0=-91.52 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs") +
  annotation_scale(location = "bl", width_hint = 0.25) +
  theme_void() +
  theme(panel.background = element_rect(fill = "aliceblue", size = 1),
         plot.background = element_blank())

inset
```
Lastly, we will add inset the inset to the main map using the `cowplot` package. We specify which is the main map using `ggdraw` and add the inset with a second function called `draw_plot`. Within `draw_plot` we define the size of the inset with `width` and `height`, as well as the location within the main map margins using `x` and `y`.

```{r warning = FALSE, message = FALSE, fig.dim = c(7, 7)}
NL.full <- ggdraw(final) + 
  draw_plot(inset, width = 0.3, height = 0.3, x = 0.25, y = 0.05)

NL.full
```

### Exporting your map

You can now export your map to use for a report, presentation, etc. One option is to use the method outlined in Chapter \@ref(export). However, we can also use the function `ggsave()`. This will allow us to name our map as well as set the file format (.PNG, .JPEG, .TIFF), size (width and height in inches), and image resolution (dpi) when exporting the map.

```{r, eval=F, echo=T}
ggsave("map_final.png", plot = NL.full, width = 7, height = 7, dpi = "screen")
```

## Additional Reading
* https://www.earthdatascience.org/courses/earth-analytics/spatial-data-r/understand-epsg-wkt-and-other-crs-definition-file-types/
* https://pro.arcgis.com/en/pro-app/2.7/help/mapping/properties/coordinate-systems-and-projections.htm
* https://cran.r-project.org/web/packages/sf/vignettes/sf1.html



