# Advanced graphing - silhouettes {#silhouettes}

```{r, echo = F}
library(rphylopic)
library(ggplot2)
```

## Introduction

You can add silhouettes to both base and ggplot2 plots using the [rphylopic](https://cran.r-project.org/web/packages/rphylopic/readme/README.html) package.

You can get free png silhouettes of various animals from [phylopic.org](http://phylopic.org). Once you install and load the `rphylopic` package, you can use the function `image_data()` to call the image's universally unique identifier (UUID) and specify the size of the image.

```{r}
squid.img <- image_data("938a7ea7-5262-4bbb-85ba-cb39181002f0", size = "512")[[1]]
```

## Adding silhouettes in ggplot

Before beginning, ensure you have installed and loaded both `ggplot2` and `rphyolopic`. Silhouettes can be added to a ggplot graphic by using the function [add_phylopic()](https://www.rdocumentation.org/packages/rphylopic/versions/0.3.0/topics/add_phylopic) as another layer.

```{r}
squid.img <- image_data("938a7ea7-5262-4bbb-85ba-cb39181002f0", size = "512")[[1]]

data = data.frame(x = 1:10, y = c(2, 9, 4, 6, 5, 8, 2, 3, 7, 1))

ggplot() + 
  geom_point(data = data, aes(x = x, y = y)) + 
  add_phylopic(squid.img)
```

The `add_phylopic()` function will automatically centre your image in the background of the plot. You can customize the location, size, colour and transparency (using alpha) of the image within the function. 

```{r}
ggplot() + 
  geom_point(data = data, aes(x = x, y = y)) + 
  add_phylopic(squid.img, alpha = 0.2, x = 8, y = 8, ysize = 3, color = "blue")
```

Troubleshooting tip: The following code from [here](https://scrogster.wordpress.com/2014/06/02/adding-phylopic-org-silhouettes-to-r-plots/), will work if the rphylopic package isn’t working well:

```{r}
# You need to install these packages first
library(png)
library(RCurl)
library(ggplot2)
library(rphylopic)

# go to phylopic.org and search for your image. Click on the image you want (twice) # and the url should be similar to as below:

squidurl<-"http://phylopic.org/assets/images/submissions/938a7ea7-5262-4bbb-85ba-cb39181002f0.512.png"
kiwiurl<-"http://phylopic.org/assets/images/submissions/ad7d0f80-889e-4dba-a89d-c864f2f48184.512.png"

# Read in the PNG file:
squid <-  readPNG(getURLContent(squidurl))
kiwi <-  readPNG(getURLContent(kiwiurl))

# Some fake data to make a graph with
data = data.frame(x = seq(1,10), y = sample(1:10))

g=ggplot(data, aes(x=x,y=y))+
  geom_point()

# Note that add_phylopic is added as a layer to the ggplot()
g + add_phylopic(squid,alpha=.5,x=5,y=7,ysize=7) + add_phylopic(kiwi,1,3,4,2)
```

Optional. View visualizations that can be produced with some advanced geometries: [ggridges](https://cran.r-project.org/web/packages/ggridges/vignettes/gallery.html), [ggviolin](https://rpkgs.datanovia.com/ggpubr/reference/ggviolin.html), [ggmarginal](https://r-graph-gallery.com/277-marginal-histogram-for-ggplot2.html). The key to getting any of these advanced visualizations to work is to have the right information formatted correctly in a dataframe. Of course you should also consider that visualization best communicates the important information.
