# R markdown {#rmarkdown}

## Introduction

In general, you’d better leave at least one empty line between adjacent but different elements, e.g., a header and a paragraph.

## Syntax

## Headers

Section headers can be written after a number of pound signs, e.g.,

```{r}
# First-level header

## Second-level header

### Third-level header

```

If you do not want a certain heading to be numbered, you can add {-} or {.unnumbered} after the heading, e.g.,

```{r}
# Preface {-}
```

## Lists

Unordered list items start with *, -, or +, and you can nest one list within another list by indenting the sub-list, e.g.,

```
- one item
- one item
- one item
    - one more item
    - one more item
    - one more item

```

The output is:

- one item
- one item
- one item
  - one more item
  - one more item
  - one more item

Ordered list items start with numbers (you can also nest lists within lists), e.g.,
```
1. the first item
2. the second item
3. the third item
    - one unordered item
    - one unordered item
```

The output does not look too much different with the Markdown source:

1. the first item
2. the second item
3. the third item
  - one unordered item
  - one unordered item
